Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-02 paper/paper.tex
comitting

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\usepackage{algorithmic}

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

\begin{abstract}
We  consider  two  ways  to  reduce the  memory  requirement  of  lazy
functional  programming  languages. The  first  is  to modify  garbage
collectors  to  preserve  only  {\em  live} data---a  subset  of  {\em
reachable data}.  This results in an increase in the garbage reclaimed
and  consequently in  fewer  garbage collections.   The  second is  to
eagerly evaluate closures  that are guaranteed to be  evaluated in the
future.

The  main contribution  of  this  paper is  an  uniform framework  for
obtaining the information for the optimizations mentioned above---{\em
liveness}  and {\em strictness}  analysis. Using  a central  notion of
demand,  we  formulate   a  context-sensitive  liveness  analysis  for
structured  data and  prove  it  correct.  We  then present strictness
as a dual of liveness. We use a  0-CFA-like
conservative   approximation   to   annotate   each   allocation   and
function-call program point  with a finite-state automaton---which the
garbage-collector inspects to  curtail reachability during
marking. The result of the strictness analysis, being necessarily
finite,  can  be directly used to insert code to evaluate closures.

%As a result,  fewer objects are marked (albeit  with a more expensive
%marker) and then preserved (e.g.\ by a copy phase).

Experiments  confirm  the  expected  performance benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.
\end{abstract}

@@

Functional  programming  languages  depend  on  garbage  collector  to
efficiently  reclaim  unused   memory  allocated  by  programs  either
explicitly  (for  example using  constructors)  or implicitly  (during
creation    of    closures).     However,   empirical    studies    on
Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
Java~\cite{shaham02estimating}   programs  have  shown   that  garbage
collectors leave uncollected a lot of memory objects that are not live
(here {\em live}ness  means the object can potentially  be used by the
program at  a later  stage).  Due  to lack of  a suitable  analysis to
distinguish  reachable  objects  from  live objects,  current  garbage
collectors conservatively approximate the  liveness of heap objects by
their reachability from a set of memory locations called the {\em root
  set\/}. This results in unnecessary memory retention.

Programs written in languages having lazy semantics for evaluation put
an further demands on  memory as they require heap allocated
closures to be  carried from the point of  occurrence of an expression
to the point of (first) real use of that expression.
%% This   may   be   mitigated  by   strictness
%% analysis~\cite{some-refs} that  can statically determine  the patterns
%% of future usage  of an expression, thus allowing  some of the closures
%% to be  evaluated at the point  of the occurrence  of the corresponding
%% expression.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

\begin{figure*}[t!]
%%%%{\color{Myblue}
\begin{picture}(100,130)(0,-90)
%  \begin{center}
    \begin{tabular}{cc}
\begin{boxedminipage}{.5\textwidth}
      {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
	       \listb
	  \UNL{1}      (\CONS\ (\CAR\  \lista)
%          \UNL{2}          (\CAR\  \lista)
	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
          \listb))))
          \UNL{0}
	  \UNL{0} \hspace*{-.49\TAL} (\LET\ \pz  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
	  \UNL{0} (\LET\ \py  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
	  \UNL{1}   (\LET\ \pw $\leftarrow$ (\append\ \py\  \pz)\ \IN
	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ \pw)))))
	\end{uprogram}
      }} 
\end{boxedminipage}
      &
      \raisebox{-25mm}{\scalebox{.75}{
	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
      \psset{unit=1mm}
      \psset{linewidth=.3mm}
      \begin{pspicture}(0,-5)(70,60)
	%\psframe(0,0)(73,60)
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	%      fillcolor=white]{$\times$}}}
        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
		  fillcolor=white]{$\times$}}}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
	\ncline[nodesepB=-.5]{*->}{c2}{d}
	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
        %here
        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}

%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%	      fillcolor=white]{$\times$}}} %and here
	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%	      fillcolor=white]{$\times$}}} %and here

	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      \end{pspicture}}} \\
      (a) Example program.&
      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
	(b) Memory graph at $\pi$. Thick edges denote \\
        {\white (b) }live links. Traversal stops at edges marked\\
        {\white (b) }$\times$  during garbage collection.
      \end{tabular}}
    \end{tabular}
\end{picture}

@@

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure*}
We propose the  use of liveness analysis of heap  cells and strictness
analysis  of expressions  to  reduce the  memory  requirement of  lazy
functional  programming languages.   We modify  garbage collectors  to
preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
This results in an increase  in the garbage reclaimed and consequently
in fewer  garbage collections.  
Using  a central notion  of demand,  we formulate  a context-sensitive
liveness  analysis  and  prove   it  correct.   We  use  a  0-CFA-like
conservative   approximation   to   annotate   each   allocation   and
function-call program point  with a finite-state automaton---which the
garbage-collector inspects to curtail reachability during marking.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order  eager  language,  a  straightforward  extension  of  this
earlier  technique  is  not   possible  in  lazy  languages,  where  a
heap-allocated   object  may  either   be  a   evaluated  data   or  a
closure. Since liveness is brought about by evaluation of closures and
given the  uncertainty in lazy  languages regarding the  program point
where this  evaluation takes  place, this introduces  complications in
the analysis itself.  Further,  since we want to extend liveness-based
collection to closures, we  track closures  to their
points of creation, from where we obtain their liveness. Since this is
done  at  runtime,  this  requires considerable  modification  to  the
earlier GC scheme.

%% We  previously  proposed an  interprocedural  liveness analysis  to
%% improve   garbage   collection   first   order  eager   subset   of
%% Scheme~\cite{asati14lgc}.   The current paper  extends the  work to
%% handle lazy evaluation semantics for the same language.

Experiments  confirm  the expected  performance  benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections,  a  (some quantification)  decrease  in  the memory  size
required to run programs,  and reduced overall garbage collection time
for a majority of programs.

%------------------------------------------------------------%
\subsubsection{Motivating Example}
\label{sec:motiv}

Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
$\pi$ of an expression $e$  denotes a program point.  During execution
of  the program, it  represents the  instant of  time just  before the
evaluation  of  $e$.  
We  view the  heap  as a graph.  
Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
containing values.  Edges in the graph are {\em references} and
emanate from {\em variables} or fields.
Variable and field values may also be atomic values ($\NIL$, integers etc.)
While it is convenient to box these in diagrams, our presented
analysis treats them as non-heap values.

Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
The edges shown  by thick arrows are those  which are made live by the
program. In addition,  assuming that
the value of any reachable part of the program result may be explored
or printed, the edges marked by thick dashed arrows are also live.
A cell  is  marked and preserved  during  garbage collection,  only  if it  is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if  a garbage collection takes
% place   at    the   time   represented   by   the    heap   shown   in
at $\pi$ with the    heap   shown   in
Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
will be marked and preserved.

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational} describes the operational semantics of
% the language.
Liveness analysis is  described in Section~\ref{sec:liveness} followed
by  a  sketch  of  a  correctness proof  relative  to  a  non-standard
semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
as   finite-state  automata.    Section~\ref{sec:experiments}  reports
experimental  results  and Section~\ref{sec:lgc-always-better}  proves
that a liveness based collector  can never do more garbage collections
than a reachability based collector.
% and Section~\ref{sec:conclusion} concludes the paper.

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\usepackage{algorithmic}

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

\begin{abstract}
We consider  the problem of reducing  the memory required  to run lazy
first-order functional programs.  To  do this, we analyze programs for
liveness of heap  allocated data, i.e. data that  can possibly be used
in the future.  The accompanying garbage collector is  modified to use
the result of the analysis to preserve only {\em live} data---a subset
of {\em reachable data}---during  garbage collection.  This results in
an  increase in  the garbage  reclaimed and  consequently  reduces the
memory footprints of programs.   While this technique has already been
shown to yield benefits  for eager first-order languages, the presence
of  closures   at  run-time  poses  additional   challenges  for  lazy
languages.   Closures require  changes  both in  the earlier  liveness
analysis for heap allocated memory  and also in the garbage collection
scheme required to reclaim memory that is not live.

We  formulate  a context-sensitive  liveness  analysis  and using  the
results annotate each potential allocation point in the program with a
set     of    deterministic    finite-state     automata(DFA).     The
garbage-collector  inspects  the DFA  to  curtail reachability  during
marking.  As a result, fewer  objects are marked (with a possibly more
expensive  marker)  and  preserved  by the  copy  phase.   Experiments
confirm  that  liveness-based  collection  results in  a  increase  in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections. On  our benchmark programs, we obtain  a decrease between
x\% and y\% in the memory size required to run programs, when compared
to reachability-based garbage collection.
\end{abstract}

@@

Functional programs extensively use dynamically allocated memory.  The
allocation  is either  explicit (for  example, using  constructors) or
implicit  (during creation  of  closures).  Programs  written in  lazy
functional languages put additional  demands on memory as they require
closures to  be carried  from the  point of creation  to the  point of
evaluation.

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}   and,   more   importantly,   on
Haskell~\cite{rojemo96lag}
%and Java~\cite{shaham02estimating}
programs have shown that  garbage collectors leave uncollected a large
number of  memory objects that are  reachable but not  live (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

\begin{figure*}[t!]
%%%%{\color{Myblue}
\begin{picture}(100,130)(0,-90)
%  \begin{center}
    \begin{tabular}{cc}
\begin{boxedminipage}{.42\textwidth}
      {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
          \UFL\
\UNL{0} (\DEFINE\ (\length\  \pl)
	  \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
          (\PRIM\ 1\ (\CAR\  \pl))))
           \UNL{0}
	   \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	  \UNL{1}  (\SIF~(\NULLQ \ \lista)
	       \listb
	  \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
(\CDR\  \lista)
          \listb))))
          \UNL{0}
          \UNL{0} (\LET\ \px\
          $\leftarrow$\ (\CONS\ $5$
%          \UNL{1}\hspace*{1.5cm}
          (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	  \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	  \UNL{2}
          (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
          \UNL{3} (\SIF~(\NULLQ~(\CAR~(\CDR~\pz)))~$0$~$\pi$:\
(\length\ \pz))))))
	\end{uprogram}
      }}
\end{boxedminipage}
&
\begin{minipage}{.45\textwidth}

\hspace*{2cm}{\epsfig{file=motiv-example.eps, height=4cm}}

\end{minipage}

\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       &
%%       \raisebox{-25mm}{\scalebox{.75}{
%% 	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
%%       \psset{unit=1mm}
%%       \psset{linewidth=.3mm}
%%       \begin{pspicture}(0,-5)(70,60)
%% 	%\psframe(0,0)(73,60)
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
%% 	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
%% %
%%      \putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
%%
%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
%%      fillstyle=solid,
%% 	%      fillcolor=white]{$\times$}}}
%%
%%  \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox
%%{.63}{\nilfigure}}}
%% %	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
%% 	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
%%
%% \putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
%% 	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
%%
%% \aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
%%fillstyle=solid,
%% 		  fillcolor=white]{$\times$}}}
%% 	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
%% 	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
%% 	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
%% 	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
%% 	\ncline[nodesepB=-.5]{*->}{c2}{d}
%% 	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
%% 	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
%%
%% \aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
%% fillstyle=solid,
%% 	      fillcolor=white]{$\times$}}}
%% 	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
%%
%% \nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
%% 	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
%%
%% \aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,
%% fillstyle=solid,
%% 	      fillcolor=white]{$\times$}}}
%%
%% \putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
%%
%%\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
%%
%%\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
%% 	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
%% 	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
%% %
%%\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
%% 	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
%%
%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
%%         %here
%%
%%\putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
%%
%%\putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{
%%\scalebox{.63}{\nilfigure}}}

%% %
%%\aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,
%%fillstyle=solid,
%% %	      fillcolor=white]{$\times$}}} %and here
%% 	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
%%
%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h}
%% %
%%\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,
%%fillstyle=solid,
%% %	      fillcolor=white]{$\times$}}} %and here

%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
%% 	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
%% 	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
%%
%%\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%% 	      fillcolor=white]{$\times$}}}
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
%% %
%%\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
%% 	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
%%
%%\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
%% %	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
        %%
%%       \end{pspicture}}} \\
%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

      (a) Example program.&\hspace*{2cm}
      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
	(b) Memory graph at $\pi$. Thick edges denote  \\
        {\white (b) }live links. Traversal stops at edges marked\\
        {\white (b) }$\times$  during garbage collection.
      \end{tabular}}
    \end{tabular}
\end{picture}

@@

\caption{Example Program and its Memory Graph.{\color{red}Amey will combine this with the figure for syntax}}\label{fig:mot-example}
\end{figure*}

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The analysis  has two  parts.  First,  each  function is
summarized in a context-sensitive manner as a {\em demand transformer}
that transforms a {\em symbolic} demand  on its body to demands on its
arguments.  This summary  is then used to step  through function calls
during analysis.   Second, the concrete  demand on a function  body is
obtained through a 0-CFA-like conservative approximation that combines
the  demands on  all the  calls to  the function.   The result  of the
analysis  is the annotation  of each  program point  with finite-state
automata capturing  the liveness of variables at  the point. Depending
on the program point where  garbage collection has been triggered, the
garbage-collector consults  a set of automata  to curtail reachability
during marking.  This results in  an increase in the garbage reclaimed
and consequently in fewer garbage collections.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order {\em  eager} language, a straightforward  extension of the
earlier  technique  is  not  possible  for  lazy  languages,  where  a
heap-allocated  object may either  be an  unevaluated expression  or a
closure. Since  data is  made live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined,  laziness complicates liveness
analysis itself.  In  addition, apart from evaluated data,  we want to
extend  liveness-based collection also  to closures---closures with no
future use should also  be collected.  This involves tracking closures
to  their points of  creation at  runtime from  where we  obtain their
liveness.  This  requires significant  modification to the  earlier GC
scheme.

{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
single
  generation  copying  collector   confirm  the  expected  performance
  benefits.   Liveness-based  collection  results  in an  increase  in
  garbage reclaimed.   As a  consequence, there is  a decrease  in the
  number of  collections and a  (some quantification) decrease  in the
  memory size required  to run programs. Besides there  is a reduction
  in the overall execution time for some programs.}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}

Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at the program point $\pi$, i.e. just before the  evaluation
of $(\length\ \pz)$.  The heap is represented by a graph in which
a node  either represents atomic values ($\NIL$,  integers etc.), or
a \CONS\  cell  containing $\CAR$  and  $\CDR$  fields, or a chunk of
memory representing an unevaluated
value  also called a  {\em closure}  (represented by  hatched boxes).
Edges in the graph are  {\em references} and emanate from variables or
fields.  The figure shows the lists \px\ and \pz\ partially evaluated.
The evaluation was triggered by the test (\NULLQ~(\CAR~(\CDR~\pz))) in
the \SIF\ expression.

The edges shown by thick arrows  are those which are live at $\pi$.  A
cell is marked and preserved  during garbage collection only if it is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if a garbage  collection takes place at
$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), only the
cells referenced by $\pz$,  $(\CDR~ \pz)$ and (the cells constituting)
the closure  referenced by $(\CDR~(\CDR~\pz))$  need to be  marked and
preserved.  In  contrast, a  reachability-based  collector would  have
preserved all cells.

{\color {Myblue}Rewrite at the end: The  specific contributions of the
paper
  are: The formulation of a  liveness analysis for a lazy language and
  the proof of its correctness, the undecidability of the grammar that
  results from the analysis and  its approximation by DFAs, the design
  of a  garbage collector that can  use the DFAs to  retain live cells
  that  also include  closures, and  empirical results  that  show the
  effectiveness of liveness-based garbage collection.}

\subsection{Organization of the paper}
